My Publication Title --- Single Author

### 2025-07-17 paper/quarto_aaai/paper.tex

Today's prominence of artificial intelligence (AI) has largely been
driven by the success of representation learning with large degrees of
freedom: instead of relying on features and rules hand-crafted by
humans, modern machine learning (ML) models are tasked with learning
highly complex representations directly from the data, guided by narrow
objectives such as predictive accuracy \citep{goodfellow2016deep}. These
models tend to be so complex and opaque that humans cannot easily
interpret their decision-making.

Counterfactual explanations (CE) have become an integral part of the
broader explainable AI (XAI) toolkit \citep{molnar2022interpretable}
that can be employed to make sense of this complexity. Originally
proposed by \citet{wachter2017counterfactual}, CEs prescribe minimal
changes for factual inputs that, if implemented, would prompt some
fitted model to produce an alternative, more desirable output
(Figure~\ref{fig-poc}). This is useful and necessary to not only
understand how opaque models make predictions, but also to provide
recourse to individuals subjected to them: a retail bank, for example,
could use CE to provide meaningful feedback to unsuccessful loan
applicants that were rejected based on an opaque AI decision-making
system (ADM).

For feedback to be meaningful, counterfactual explanations need to
fulfill certain desiderata
\citep{verma2020counterfactual, karimi2020survey}: they should be
faithful to the model \citep{altmeyer2024faithful}, plausible
\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
Plausibility is typically understood as counterfactuals being
\emph{in-domain}: unsuccessful loan applicants that implement recourse
should end up with a credit profile that is genuinely similar to that of
individuals who have successfully repaid their loans in the past.
Actionable explanations comply with practical constraints: a young,
unsuccessful loan applicant cannot increase their age in an instance.

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because a narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore follow the
paradigm that models---as opposed to explanation methods---should be
held accountable for plausible and actionable explanations. While
previous work has shown that at least plausibility can be indirectly
achieved through existing techniques aimed at models' generative
capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful CE with
ground-truth data while protecting immutable features, and capitalize on
methodological links between CE and AE by penalizing the model's
adversarial loss on interim (\emph{nascent}) counterfactuals. To the
best of our knowledge, CT represents the first venture in this direction
with promising results, both empirical and theoretical.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) implausibility of CE decreases by 16\% on average; (2) the
cost of reaching valid counterfactuals with protected features decreases
by 15\%; and (3) robust accuracy increases for all datasets by up to 42
percentage points. Finally, in Section~\ref{sec-discussion}, we discuss
open challenges and conclude.

@@

\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
classifiers trained under different regimes on synthetic data: (a)
conventional training, all mutable; (b) CT, all mutable; (c)
conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
linear decision boundary is shown in green along with training data
colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
(orange). Stars indicate counterfactuals in the target class. Synthetic
features are annotated as \emph{debt} and \emph{age} for illustrative
purposes.}

@@

To make the desiderata for our framework more concrete, we follow
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to `explanatory capacity' simply as `explainability' in what
follows (see Def. \ref{def-explainability} below for a more formal
exposition).

@@

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. To do so, the authors
first propose a way to generate counterfactuals that are maximally
faithful to the model in that they are consistent with what the model
has learned about the underlying data. Formally, they rely on tools from
energy-based modelling \citep{teh2003energy} to minimize the divergence
between the distribution of counterfactuals and the conditional
posterior over inputs learned by the model. Their counterfactual
explainer, \emph{ECCCo}, yields plausible explanations if and only if
the underlying model has learned representations that align with them.
The authors find that both deep ensembles
\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\citep{grathwohl2020your}---a form of constrastive learning---tend to do
well in this regard.

@@

\emph{Interventions on features have implications for fairness.} We
provide a tool that allows practitioners to modify the sensitivity of a
model with respect to certain features. Model owners can use our
solution to support the fair and equitable treatment of decision
subjects, but they could also misuse it by enforcing explanations based
on features that are more difficult to modify by some (group of)
individuals. When used irresponsibly, CT could result in an unfairly
assigned burden of recourse \citep{sharma2020certifai}, threatening the
equality of opportunity \citep{bell2024fairness}. Additionally, CT
requires mutability constraints for the features considered by the
model. Even if all immutable features are protected, there may exist
proxies that are mutable, and hence should not be protected, but
preserve sufficient information about the principals to hinder these
protections. Deciding on actionability is still a major open challenge
in the AR literature \citep{venkatasubramanian2020philosophical}
impacting the capacity of CT to fulfill its intended goal.

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of learned representations but they also increase
the number of computations. As our codebase is not performance
optimized, grids of 270 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

We also highlight three important directions for future research.
Firstly, it is an interesting challenge to extend CT beyond
classification settings. Our formulation relies on the distinction
between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
and \(y^{+}\) can be arbitrarily defined, CT requires the output space
\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
where the change in outcome cannot be readily discretized. Focus on
classification is a common choice in research on CEs and AR. Other
settings have attracted some interest, e.g., regression
\citep{spooner2021counterfactual}, but there is little consensus how to
robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities. This problem has
been recognized for JEMs \citep{grathwohl2020your} and even though we
depart from the SGLD sampling, we still encounter variability in
outcomes. CT is exposed to two potential sources of instabilities: (1)
the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
Equation~\ref{eq-div}, and (2) the underlying explainers. We find
several promising ways to mitigate this problem: regularizing energy
(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
during each epoch, and including only mature counterfactuals in
\(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future research could benefit
from more sophisticated approaches.

To conclude, state-of-the-art machine learning models are prone to
learning complex representations that cannot be interpreted by humans.
Existing explainability solutions cannot guarantee that explanations
agree with the model's learned representation of data. As a step towards
addressing this challenge, we introduced counterfactual training, a
novel training regime that incentivizes highly-explainable models. Our
approach leads to explanations that are both plausible (compliant with
the underlying data-generating process) and actionable (compliant with
user-specified mutability constraints), and thus meaningful to their
recipients. Through extensive experiments, we demonstrated that CT
satisfies its objective while promoting robustness and preserving the
predictive performance of models. It can also be used to fine-tune
conventionally-trained models and achieve similar gains. Lastly, our
work highlights the value of simultaneously improving models and their
explanations.

>>>>

Today's prominence of artificial intelligence (AI) has largely been
driven by the success of representation learning with large degrees of
freedom: instead of relying on features and rules hand-crafted by
humans, modern machine learning (ML) models are tasked with learning
highly complex representations directly from the data, guided by narrow
objectives such as predictive accuracy \citep{goodfellow2016deep}. These
models tend to be so complex that humans cannot easily interpret their
decision-making.

Counterfactual explanations (CE) have become a key part of the broader
explainable AI (XAI) toolkit \citep{molnar2022interpretable} that can be
applied to make sense of this complexity. Originally proposed by
\citet{wachter2017counterfactual}, CEs prescribe minimal changes for
factual inputs that, if implemented, would prompt some fitted model to
produce an alternative, more desirable output (Figure~\ref{fig-poc}).
This is useful and necessary to not only understand how opaque models
make their predictions, but also to provide algorithmic recourse (AR) to
individuals subjected to them: a retail bank, for example, could use CE
to provide meaningful feedback to unsuccessful loan applicants that were
rejected based on an opaque automated decision-making (ADM) system.

For such feedback to be meaningful, counterfactual explanations need to
fulfill certain desiderata
\citep{verma2020counterfactual, karimi2020survey}---they should be
faithful to the model \citep{altmeyer2024faithful}, plausible
\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
Plausibility is typically understood as counterfactuals being
\emph{in-domain}: unsuccessful loan applicants that implement the
provided recourse should end up with credit profiles that are genuinely
similar to that of individuals who have successfully repaid their loans
in the past. Actionable explanations comply with practical constraints:
a young, unsuccessful loan applicant cannot increase their age in an
instant.

Existing state-of-the-art (SOTA) approaches in the field have largely
focused on designing model-agnostic CE methods that identify subsets of
counterfactuals, which comply with specific desiderata. This is
problematic, because a narrow focus on any specific desideratum can
adversely affect others: it is possible, for example, to generate
plausible counterfactuals for models that are also highly vulnerable to
implausible, possibly adversarial counterfactuals
\citep{altmeyer2024faithful}. In this work, we therefore follow the
paradigm that models---as opposed to explanation methods---should be
held accountable for explanations that are plausible and actionable.
While previous work has shown that at least plausibility can be
indirectly achieved through existing techniques aimed at models'
generative capacity, generalization and robustness
\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
we directly incorporate both plausibility and actionability in the
training objective of models to improve their overall explanatory
capacity.

Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
training regime that leverages counterfactual explanations on-the-fly to
ensure that differentiable models learn plausible and actionable
explanations for the underlying data, while at the same time also being
more robust to adversarial examples (AE). Figure~\ref{fig-poc}
illustrates the outcome of CT compared to conventionally trained models:
in panel (a), faithful and valid counterfactuals end up near the
decision boundary forming a clearly distinguishable cluster in the
target class (orange); in panel (b), CT is applied to the same
underlying linear classifier architecture resulting in much more
plausible counterfactuals; in panel (c), the classifier is again trained
conventionally and we have introduced a mutability constraint on the
\emph{age} feature at test time---counterfactuals are valid but the
classifier is roughly equally sensitive to both features; by contrast,
the decision boundary in panel (c) has titled, making the model trained
with CT relatively less sensitive to the immutable \emph{age} feature.
To achieve these outcomes, CT draws inspiration from the literature on
contrastive and robust learning: we contrast faithful (\emph{mature})
CEs with ground-truth data while protecting immutable features, and
capitalize on methodological links between CE and AE by penalizing the
model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
the best of our knowledge, CT represents the first venture in this
direction with promising empirical and theoretical results.

The remainder of this manuscript is structured as follows.
Section~\ref{sec-lit} presents related work, focusing on the links to
contrastive and robust learning. Then follow our two principal
contributions. In Section~\ref{sec-method}, we introduce our
methodological framework and show theoretically that it can be employed
to enforce global actionability constraints. In our experiments
(Section~\ref{sec-experiments}), we find that thanks to counterfactual
training, (1) the implausibility of CEs decreases by 16\% on average;
(2) the cost of reaching valid counterfactuals with protected features
decreases by 15\%; and (3) the robust accuracy increases by up to 42
percentage points. Finally, we discuss open challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion}.

@@

\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
classifiers trained under different regimes on synthetic data: (a)
conventional training, all mutable; (b) CT, all mutable; (c)
conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
linear decision boundary is shown in green along with training data
colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
(orange). Stars indicate counterfactuals in the target class. For
illustration, synthetic features are annotated as \emph{debt} and
\emph{age}.}

@@

To make the desiderata for our framework more concrete, we follow the
previous work in tying the explanatory capacity of models to the quality
of CEs that can be generated for them
\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
we refer to ``explanatory capacity'' as ``explainability'' in the rest
of this manuscript (see Def. \ref{def-explainability} below for a more
formal exposition).

@@

In a closely related work, \citet{altmeyer2024faithful} show that model
averaging and, in particular, contrastive model objectives can produce
more explainable and hence trustworthy models. The authors propose a way
to generate counterfactuals that are maximally faithful to the model in
that they are consistent with what the model has learned about the
underlying data. Formally, they rely on tools from energy-based
modelling \citep{teh2003energy} to minimize the divergence between the
distribution of counterfactuals and the conditional posterior over
inputs learned by the model. Their counterfactual explainer,
\emph{ECCCo}, yields plausible explanations if and only if the
underlying model has learned representations that align with them. The
authors find that both deep ensembles \citep{lakshminarayanan2016simple}
and joint energy-based models (JEMs) \citep{grathwohl2020your}, a form
of constrastive learning, tend to do well in this regard.

@@

\emph{Interventions on features have implications for fairness.} We
provide a method to modify the sensitivity of a model to certain
features, which can be misused by enforcing explanations based on
features that are more difficult to modify by a (group of) decision
subjects. Such abuse could result in an unfairly assigned burden of
recourse \citep{sharma2020certifai}, threatening the equality of
opportunity \citep{bell2024fairness}. Also, even if all immutable
features are protected, there may exist proxies that are theoretically
mutable, but preserve sufficient information about the principals to
hinder these protections. Indeed, deciding on the actionability of
features remains a major open challenge in the AR literature
\citep{venkatasubramanian2020philosophical}.

\emph{CT increases the training times.} Like adversarial training, CT is
more resource-intensive than conventional regimes. Higher numbers of CEs
improve the quality of the learned representations but they also
increase the number of computations. As our codebase is not performance
optimized, grids of around 300 settings for the largest datasets in our
experiments took up to four hours using 34 2GB CPUs (see supplementary
appendix). Other than optimization, three factors mitigate this effect:
(1) CT yields itself to parallel execution; (2) it amortizes the cost of
CEs for the training samples; and (3) it can be used to fine-tune
conventionally-trained models.

We also highlight three key directions for future research. Firstly, it
is an interesting challenge to extend CT beyond classification settings.
Our formulation relies on the distinction between non-target class(es)
and target class(es), requiring the output space to be discrete. Thus,
it does not apply to ML tasks where the change in outcome cannot be
readily discretized. Focus on classification is a common choice in
research on CEs and AR; other settings have attracted some interest,
e.g., regression \citep{spooner2021counterfactual}, but there is little
consensus how to robustly extend the notion of CEs.\\
Secondly, CT is susceptible to training instabilities, as also
recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
to two sources of instabilities: (1) the energy-based contrastive
divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
(2) the underlying explainers. We find several promising ways to
mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
generating sufficiently many counterfactuals during each epoch, and
including only mature counterfactuals in \(\text{div}(\cdot)\).\\
Finally, we believe that it is possible to considerably improve
hyperparameter selection procedures, and thus performance. We have
relied exclusively on grid searches, but future work could benefit from
more sophisticated approaches.

\section{Conclusion}\label{sec-conclusion}

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Existing
explainability solutions cannot guarantee that explanations agree with
these learned representation. As a step towards addressing this
challenge, we introduce counterfactual training, a novel training regime
that incentivizes highly-explainable models. Through extensive
experiments, we demonstrate that CT satisfies this objective while
promoting robustness and preserving the predictive performance of
models. Explanations generated from CT-based models are both more
plausible (compliant with the underlying data-generating process) and
more actionable (compliant with user-specified mutability constraints),
and thus meaningful to their recipients. We also show that CT can be
used to fine-tune conventionally-trained models and achieve similar
gains. Lastly, our work highlights the value of simultaneously improving
models and their explanations.
